---
title: "Project 3"
author: "Md. Simon Chowdhury, Inna Yedzinovich, Zaneta Paulusova, Md Asaduzzaman, Md. Asadul"
date: "2024-10-30"
output: html_document
---

```{r setup, include=FALSE}
# Load necessary libraries
library(tidyverse)

# Set options
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and Exploring the Data
Let's load the dataset and take a quick look at the first few rows to understand its structure.
```{r}
# Load data
data <- read_csv("https://raw.githubusercontent.com/simonchy/DATA607/refs/heads/main/week%208/Cleaned_Augmented_Data_Science_Skills.csv")

# Display the first few rows of the data
head(data)

```

## Data Cleaning
We’ll clean up the data by renaming columns and removing rows with missing values in skill columns.

```{r}
# Standardize column names to avoid spaces and special characters
colnames(data) <- make.names(colnames(data))

# Select only the columns with the top skills
skills_data <- data %>%
  select(Name..1.most.most.valuable.data.science.skill,
         Name..2.most.most.valuable.data.science.skill,
         Name..3.most.most.valuable.data.science.skill,
         Name..4.most.most.valuable.data.science.skill,
         Name..5.most.most.valuable.data.science.skill) %>%
  pivot_longer(cols = everything(),
               names_to = "skill_rank",
               values_to = "skill") %>%
  filter(!is.na(skill)) # Remove rows with NA skills

# Check cleaned and reshaped data
head(skills_data)

```

## Analysis: Most Valued Data Science Skills
Let's calculate the frequency of each skill to determine which are the most valued.
```{r}
# Count the occurrences of each skill
skill_counts <- skills_data %>%
  group_by(skill) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Display the top skills
head(skill_counts, 10)

```

## Visualization
Finally, we visualize the most valued data science skills using a bar plot.
```{r}
# Plot the top 10 most valued skills
library(ggplot2)

ggplot(skill_counts[1:10, ], aes(x = reorder(skill, count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Valued Data Science Skills",
       x = "Skill",
       y = "Count") +
  theme_minimal()

```

## Conclusion
Based on our analysis, the top skills identified in the dataset are:
```{r}
# Display top skills as a list
skill_counts[1:10, ]

```

These skills represent the most valued abilities in the data science field according to the responses provided and Resourcefulness is in top as most valued ability in the data science field.
