---
title: "Week 11 Assignment"
author: "MD SIMON CHOWDHURY"
date: "2024-11-16"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)  # To read Excel files
library(tidyr)   # For reshaping data
library(dplyr)   # For data manipulation
library(openxlsx) # To write Excel files
```

## Introduction

In this analysis, we will implement the **Global Baseline Estimate** algorithm to predict movie ratings. The data is loaded from an Excel file, processed to calculate global averages, user biases, and item biases, and then used to predict missing ratings.

## Steps Overview

1. Load the ratings data from an Excel file.
2. Reshape the data to a suitable format for analysis.
3. Compute the global average rating.
4. Calculate user and movie-specific biases.
5. Predict missing ratings using the formula:  
   \[
   P_{u,i} = \mu + b_u + b_i
   \]  
   Where:
   - \(\mu\): Global average rating.
   - \(b_u\): User bias.
   - \(b_i\): Item (movie) bias.

## Import the Ratings Data
We will read the ratings data from the movie_ratings.xlsx file.

```{r}
# Specify url
url <- "https://raw.githubusercontent.com/simonchy/DATA607/refs/heads/main/week%2011/MovieRatings.xlsx"

# Create a temporary file path
temp_file <- tempfile(fileext = ".xlsx")

# Download the file to the temporary location
download.file(url, destfile = temp_file, mode = "wb")

# Read the Excel file
ratings <- read_excel(temp_file)

# Display the imported data
print(ratings)

```

## Reshape the Data
To make calculations easier, we'll transform the data into a long format.

```{r}
# Reshape the data into a long format
ratings_long <- ratings %>%
  pivot_longer(-Critic, names_to = "Movie", values_to = "Rating") %>%
  drop_na()

# Display the reshaped data
head(ratings_long)

```

## Calculate Global Baseline

### Global Average (μ)
First, we calculate the global average rating across all critics and movies.

```{r}
global_mean <- mean(ratings_long$Rating)
cat("Global Average (µ):", global_mean, "\n")
```

### User Bias \( b_u \)
Next, we calculate how much each user's ratings deviate from the global average.

```{r}
user_bias <- ratings_long %>%
  group_by(Critic) %>%
  summarize(b_u = mean(Rating) - global_mean)

# Display user biases
print(user_bias)

```

### Item Bias (b_i)
We also calculate how much each movie's ratings deviate from the global average.

```{r}
item_bias <- ratings_long %>%
  group_by(Movie) %>%
  summarize(b_i = mean(Rating) - global_mean)

# Display item biases
print(item_bias)

```

## Predict Missing Ratings
Using the global average, user bias, and item bias, we compute predicted ratings.

```{r}
# Merge biases with the original data
ratings_with_bias <- ratings_long %>%
  left_join(user_bias, by = "Critic") %>%
  left_join(item_bias, by = "Movie") %>%
  mutate(Prediction = global_mean + b_u + b_i)

# Display predictions
print(ratings_with_bias)

```

## Save the Results
The predictions are saved into a new Excel file named results_ratings.xlsx.

```{r}
# Save predictions to Excel
output_file <- "results_ratings.xlsx"
write.xlsx(ratings_with_bias, output_file, append= FALSE)

cat("Predictions saved to:", output_file, "\n")

```


## Conclusion
This analysis implemented the Global Baseline Estimate algorithm to predict missing ratings.
