---
title: "Movie Ratings Analysis"
author: "MD SIMON CHOWDHURY"
date: "2024-09-15"
output: html_document
---

## Introduction

In this week, we gather movie ratings from several individuals, store them in a normalized MySQL database, transfer the data into R, handle missing data, and explore options for enhancing security and standardizing ratings.

```{r setup}
# Load required libraries
library(DBI)
library(RMySQL)

# Connect to MySQL database on localhost
conn <- dbConnect(RMySQL::MySQL(),
                  dbname = "movies_poll",
                  host = "localhost",
                  port = 3306,
                  user = "root",
                  password = Sys.getenv("MYSQL_PASSWORD"))

# Query data
movie_data <- dbGetQuery(conn, "SELECT * FROM ratings")

```

```{r}
# View the data
print(movie_data)

```

## Handling Missing Data
Several ratings are missing (NA values). I decided to impute the missing ratings by filling them with the average rating for that movie. I used the dplyr package to replace missing ratings (NA) with the average rating for the respective movie.

```{r}
# Load necessary library
library(dplyr)

# Replace NA values with the mean rating for each movie
movie_data_imputed <- movie_data %>%
  group_by(movie_id) %>%
  mutate(rating = ifelse(is.na(rating), mean(rating, na.rm = TRUE), rating)) %>%
  ungroup()

# View imputed data
print(movie_data_imputed)

```

## Standardizing Ratings
To make ratings comparable across users, I standardized the ratings using z-scores. By standardizing the ratings into z-scores, we ensure that each user's ratings are centered around their personal average, making comparisons fairer.

```{r}
# Standardize ratings by calculating z-scores
movie_data_standardized <- movie_data %>%
  group_by(person_id) %>%
  mutate(z_rating = scale(rating, center = TRUE, scale = TRUE)) %>%
  ungroup()

# View standardized ratings
print(movie_data_standardized)

```
## Conclusion

In conclusion, we collected movie ratings from five individuals and stored them in a normalized MySQL database. The ratings were transferred to R, where I handled missing data using mean imputation. The ratings were then standardized using z-scores to allow fair comparisons across different users. By using environment variables to manage sensitive information like database credentials, I improved the security of the database.

The average ratings for each movie were calculated, and the best-rated movie was identified.

```{r}
# Calculate average ratings for each movie
avg_ratings <- movie_data_imputed %>%
  group_by(movie_id) %>%
  summarise(avg_rating = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(avg_rating))

# Get the movie name for the highest-rated movie
best_movie_id <- avg_ratings$movie_id[1]
best_movie_name <- dbGetQuery(conn, paste("SELECT movie_name FROM movies WHERE movie_id =", best_movie_id))

# Display the best-rated movie
cat("The best-rated movie is:", best_movie_name$movie_name, "with an average rating of", avg_ratings$avg_rating[1], "\n")
# Close the connection
dbDisconnect(conn)

```
