---
title: "Analyzing Traffic Patterns and Predicting Congestion"
author: "MD Asadul Islam & Md Simon Chowdhury"
date: "`r Sys.Date()`"
output: slidy_presentation
---

# Introduction

- **Urban Traffic Congestion**:
  - Major issue in NYC affecting travel efficiency and quality of life.
  - Project goals:
    1. Analyze traffic data for patterns.
    2. Predict congestion hotspots.
    3. Offer actionable insights for mitigation.

- **Dataset Overview**:
  - Fields: `RequestID`, `Boro`, `Vol`, `Direction`, `WktGeom`.
  - Focus on identifying:
    - Peak traffic times.
    - High-traffic boroughs.
    - Specific congestion hotspots.

---

# Data Preparation

### Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)  # Data manipulation
library(lubridate)  # Date handling
library(ggplot2)    # Visualization
library(sf)         # Geospatial data
library(DT)         # Interactive tables
```

### Loading and Inspecting Data

```{r load_data, echo=FALSE}
traffic_data <- read.csv("C:/Users/unnia/Downloads/DAta607/final project/Automated_Traffic_Volume_Counts_20241207.csv")
head(traffic_data)
str(traffic_data)
```

**Initial Observations**:
- Rows: `r nrow(traffic_data)`
- Columns: `r ncol(traffic_data)`

---

# Data Cleaning

### Key Steps

```{r cleaning_steps}
traffic_data <- traffic_data %>%
  mutate(
    Date = make_date(Yr, M, D),
    Time = sprintf("%02d:%02d:%02d", HH, MM, 0),
    DateTime = as.POSIXct(paste(Date, Time), format="%Y-%m-%d %H:%M:%S")
  ) %>%
  filter(Vol >= 0) %>%
  distinct()

# Check for missing values in each column
missing_values <- colSums(is.na(traffic_data))

traffic_data$Vol[traffic_data$Vol < 0] <- NA
traffic_data$Vol[is.na(traffic_data$Vol)] <- median(traffic_data$Vol, na.rm = TRUE)

```

- **Negative Volumes** replaced with `NA`.
- **Duplicates Removed**.
- **Date-Time Combined** for temporal analysis.

```{r cleaned_preview, echo=FALSE}
library(DT)
datatable(head(traffic_data))
```

---

# Exploratory Data Analysis

## Traffic Volume Trends

### Volume Over Time

```{r volume_over_time}
time_volume <- traffic_data %>%
  group_by(DateTime) %>%
  summarise(Total_Vol = sum(Vol))

ggplot(time_volume, aes(x = DateTime, y = Total_Vol)) +
  geom_line(color = "blue", size = 1) +
  labs(
    title = "Traffic Volume Over Time",
    x = "Time",
    y = "Volume"
  ) +
  theme_minimal()
```

---

## Traffic Volume by Borough

```{r volume_by_borough}
borough_volume <- traffic_data %>%
  group_by(Boro) %>%
  summarise(Total_Vol = sum(Vol))

ggplot(borough_volume, aes(x = Boro, y = Total_Vol, fill = Boro)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Total_Vol), vjust = -0.5) +
  labs(
    title = "Traffic Volume by Borough",
    x = "Borough",
    y = "Volume"
  ) +
  theme_minimal()
```

---

# Focus on Congestion Hotspots

## Pulaski Bridge Case Study

```{r pulaski_bridge}
station_data <- traffic_data %>%
  filter(street == "PULASKI BRIDGE") %>%
  mutate(Hour = HH) %>%
  group_by(Hour) %>%
  summarise(Average_Volume = mean(Vol, na.rm = TRUE))

ggplot(station_data, aes(x = Hour, y = Average_Volume)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Hourly Traffic Volume at Pulaski Bridge",
    x = "Hour of the Day",
    y = "Average Traffic Volume"
  ) +
  theme_minimal()
```

---

## Interactive Hotspot Map

```{r interactive_map}
# Filter data for valid WKT geometries
traffic_data <- traffic_data %>% filter(!is.na(WktGeom))
traffic_data <- traffic_data %>% filter(grepl("^POINT\\s\\(.+\\)$", WktGeom))

# Convert to spatial data
traffic_sf <- st_as_sf(traffic_data, wkt = "WktGeom", crs = 4326)

# Fix invalid geometries
traffic_sf <- traffic_sf %>% st_make_valid()

# Plot congestion hotspots
ggplot(data = traffic_sf) +
  geom_sf(aes(color = Vol)) +
  labs(
    title = "Congestion Hotspots",
    color = "Volume"
  ) +
  theme_minimal()
```

---

# Key Insights

- **High-Traffic Boroughs**: 
  - `r borough_volume$Boro[which.max(borough_volume$Total_Vol)]` shows the highest traffic volume.
- **Peak Congestion Hours**: Evening rush hours (4–7 PM) show the highest volumes.
- **Hotspots Identified**: Pulaski Bridge and other key streets observed.

---

# Conclusion

- In conclusion, our analysis highlights peak congestion hours, high-traffic boroughs, and significant hotspots. Integrating external factors like weather and special events could enhance this study. These findings provide actionable insights for policymakers and urban planners
- **Future Directions**:
  1. Integrate weather or event data.
  2. Build predictive models to forecast congestion.
  3. Optimize traffic management strategies based on insights.

---
