---
title: "Week 3 Assignment"
author: "MD SIMON CHOWDHURY"
date: "2024-09-21"
output: html_document
---

## 1. Normalization in R

Normalization is the process of organizing data to minimize redundancy. In this section, we demonstrate normalization by splitting a single dataframe into three related dataframes.

```{r}
# Original Data
student_data <- data.frame(
  student_id = c(1, 2, 3),
  student_name = c("Alice", "Bob", "Charlie"),
  course = c("Math", "Science", "Math"),
  instructor = c("Dr. Smith", "Dr. Jones", "Dr. Smith")
)

student_data

# Normalized Data

# Students Table
students <- data.frame(
  student_id = c(1, 2, 3),
  student_name = c("Alice", "Bob", "Charlie")
)

students

# Courses Table
courses <- data.frame(
  course_id = c(1, 2),
  course_name = c("Math", "Science"),
  instructor = c("Dr. Smith", "Dr. Jones")
)

courses

# Enrollments Table
enrollments <- data.frame(
  student_id = c(1, 2, 3),
  course_id = c(1, 2, 1)
)

enrollments
```
## 2. Character Manipulation
Using the College Majors dataset from FiveThirtyEight, we identify the majors that contain either "DATA" or "STATISTICS."

```{r}
library(dplyr)

# Load data from URL
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"
majors <- read.csv(url)

# Filter majors that contain "DATA" or "STATISTICS"
selected_majors <- majors %>%
  filter(grepl("DATA|STATISTICS", Major, ignore.case = TRUE))

selected_majors

```

## 3. Regular Expressions Descriptions
Below are explanations of what the following regular expressions will match:

1. (.)\1\1: This matches any character repeated three times consecutively, such as "aaa" or "111".
   example:
```{r}
example1 <- c("aaa", "bbb", "111", "abc")
grepl("(.)\\1\\1", example1)

```


2. "(.)(.)\2\1": This matches a four-character string where the first two characters are mirrored by the       last two, like "abba".
   example:
```{r}
example2 <- c("abba", "12321", "abcd")
grepl("(.)(.)\\2\\1", example2)

```

3. (..)\1: This matches any two-character sequence that repeats, like "abab" or "1212".
   example:
```{r}
example3 <- c("abab", "1212", "abcd")
grepl("(..)\\1", example3)

```

4. "(.).\1.\1": This matches a five-character string where the first, third, and fifth characters are the      same, like "ababa" or "x2x2x".
   example:
```{r}
example4 <- c("ababa", "x2x2x", "abcde")
grepl("(.).\\1.\\1", example4)

```

5. "(.)(.)(.).*\3\2\1": This matches any string where the first three characters are repeated in reverse       order somewhere later in the string, like "abc123cba".
   example:
```{r}
example5 <- c("abc123cba", "xyz321zyx", "abcdef")
grepl("(.)(.)(.).*\\3\\2\\1", example5)

```

## 4. Constructing Regular Expressions with Examples
Here are the regular expressions constructed to match specific patterns, along with examples:

1.Start and end with the same character:
```{r}
pattern1 <- "^(.).*\1$"
example6 <- c("level", "radar", "test", "abba")
grepl(pattern1, example6)

```

2.Contain a repeated pair of letters: 
```{r}
pattern2 <- "(..).*\\1"
example7 <- c("church", "abab", "banana", "test")
grepl(pattern2, example7)

```

3.Contain one letter repeated in at least three places:
```{r}
pattern3 <- "(.).*\\1.*\\1"
example8 <- c("eleven", "banana", "abracadabra", "test")
grepl(pattern3, example8)

```
