---
title: "Week 9 Assignment"
author: "MD SIMON CHOWDHURY"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
# Load necessary libraries
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)

```

## Introduction
In this assignment we will demonstrate how to fetch and analyze books data from the New York Times Books API. The list we're interested in is "Hardcover Fiction," and we will transform the JSON response into an R data frame.

## Fetching Data from the API

```{r}
# Function to fetch and transform NY Times Books data
get_nyt_books_data <- function(api_key, list_name = "hardcover-fiction") {
  # Construct the URL
  url <- paste0("https://api.nytimes.com/svc/books/v3/lists.json?list=", list_name, "&api-key=", api_key)
  
  # Make the GET request
  response <- GET(url, accept("application/json"))
  
if (status_code(response) == 200) {
  json_data <- content(response, as = "text", encoding = "UTF-8")
  data <- fromJSON(json_data, flatten = TRUE)
  
  # Extract book details into a separate data frame
books_details_df <- data$results %>%
  mutate(book_details = map(book_details, ~ .x %>% 
                              as.data.frame() %>% 
                              select(title, author))) %>%
  unnest(book_details)

#select and work with columns directly
books_summary <- books_details_df %>%
  select(rank, title, author, weeks_on_list)

print(books_summary)
} else {
  books_df <- NULL
  print("Failed to fetch data from the API.")
}
}


api_key <- "gYMvXAooRmMGPCBIAj8ADZAIT0pBzW97"
books_data <- get_nyt_books_data(api_key)

```

## Displaying the Data
The data fetched from the New York Times Books API has been converted into a data frame and displayed below

```{r}

head(books_data)

```

## Analysis
Now that the data is loaded, lets perform some analysis.

```{r}
# Summarize the data
summary(books_data)

# Check unique book titles
unique(books_data$title)

```

## Conclusion
This assignment shows how to interface with an external API, parse JSON data, and transform it into an R data frame for further analysis. The data can be used for various visualizations and analyses in R.

