---
title: "Poll Data Analysis"
author: "MD SIMON CHOWDHURY"
date: "2024-09-07"
output: html_document
---

## Introduction

This analysis uses polling data from FiveThirtyEight, which provides a comprehensive overview of U.S. Senate races across various states. The dataset includes information from numerous polls, covering a wide array of candidates and states. It features details such as the polling firm, state, start and end dates of the polls, and the percentage of voter support for each candidate.

The primary objective of this analysis is to clean and transform this extensive dataset to focus on key variables such as the polling firm, state, candidate names, and their respective support percentages. We will identify and visualize the top candidates based on their average polling percentages to gain insights into the current standings in various Senate races.

The data for this analysis is available at [FiveThirtyEight's Senate Polls dataset](https://projects.fivethirtyeight.com/polls-page/data/senate_polls.csv).


## Load Libraries and Data

We will start by loading the necessary libraries and the dataset.

```{r setup}
# Load necessary libraries
library(dplyr)
library(readr)
library(ggplot2)

# Load the data from a CSV file
# Replace `url` with the path to your github CSV file
url <- "https://raw.githubusercontent.com/simonchy/DATA607/refs/heads/main/week%201/senate_polls.csv"
data <- read_csv(url)
```

## Data Exploration
Let's take a look at the structure of the data and preview the first few rows.

```{r}
# View the structure of the dataset
str(data)

# Preview the first few rows
head(data)
```

## Data Cleaning and Transformation
We'll now select relevant columns, rename them for clarity, and convert date columns to the appropriate format.

```{r}
# Select relevant columns and rename them
cleaned_data <- data %>%
  select(
    Pollster = pollster,
    State = state,
    StartDate = start_date,
    EndDate = end_date,
    CandidateName = candidate_name,
    Percentage = pct
  )

# Convert StartDate and EndDate to Date format
cleaned_data <- cleaned_data %>%
  mutate(
    StartDate = as.Date(StartDate, format = "%m/%d/%y"),
    EndDate = as.Date(EndDate, format = "%m/%d/%y")
  )

# Preview the cleaned data
head(cleaned_data)

```

## Data Analysis
Let's summarize the data to see the average percentage of votes each candidate received.

```{r}
# Summarize the data by candidate
summary_data <- cleaned_data %>%
  group_by(CandidateName) %>%
  summarise(
    AvgPercentage = mean(Percentage, na.rm = TRUE),
    Count = n()
  )

# Display the summary data
summary_data

```

## Visualization 
We can visualize only the top five candidates based on their average percentage of votes using a bar chart.

```{r}
# Sort the data by average percentage in descending order
top_five_candidates <- summary_data %>%
  arrange(desc(AvgPercentage)) %>%
  head(5)

# Plot only the top five candidates
ggplot(top_five_candidates, aes(x = reorder(CandidateName, -AvgPercentage), y = AvgPercentage, fill = CandidateName)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Candidates by Average Percentage of Votes", x = "Candidate", y = "Average Percentage") +
  theme(legend.position = "none")  


```

## Conclusions

The analysis of the polling data reveals that Maria Cantwell and Charles D. Baker have nearly identical average support percentages across the polls. They are followed by Tammy Baldwin, Robert P. Casey Jr., and James Justice, who also have notable support but with slightly lower average percentages compared to the top two candidates.

This finding indicates a competitive landscape among these candidates, with Cantwell and Baker standing out as the front-runners. Future analyses could focus on tracking changes in support over time or comparing these results with other polling sources to assess consistency and identify any emerging trends.

Additionally, expanding the analysis to include other factors such as polling methodology, sample size, and geographic distribution could provide a more comprehensive understanding of the dynamics in these Senate races.
